Decoupled Fallback: A Portable Single-Pass GPU Scan

### 2025-03-01 paper.tex
Conclusion edit.

As the compute capabilities of major graphics APIs---Vulkan, Direct3D, and Metal---have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers---most notably Apple's M series and ARM---lack forward-progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

@@

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving SOL performance without requiring forward-progress guarantees (FPG). Our implementation targets portability across modern GPU hardware and software platforms. It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring seamless integration with modern graphics APIs. Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. This is because workgroups are still able to proceed after an omission. In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress. Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive. Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency
\section{Discussion}

@@

\subsection{Conclusion}
In this paper, we presented \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of reaching SOL performance under all progress models and suitable for implementation within the constraints of contemporary portable shading languages. Our approach maintains minimal data movement while ensuring robust execution across diverse GPU architectures.

>>>>

As the compute capabilities of major graphics APIs have grown, there has been increasing interest in adapting \emph{Decoupled Lookback} beyond CUDA for broader adoption in real-time graphics and GPU computing. However, portable implementation is constrained by both hardware and API limitations. In terms of hardware, some GPU schedulers---most notably Apple's M series and ARM---lack forward-progress guarantees, and subgroup sizes vary across architectures. In terms of API design, the absence of 64-bit atomic operations and explicit memory barriers further complicates implementation, as both are integral to \emph{Decoupled Lookback}.

@@

We present \emph{Decoupled Fallback}, a fully portable prefix scan capable of achieving SOL performance without requiring forward-progress guarantees (FPG). Our implementation targets portability across modern GPU hardware and software platforms. It is implemented in the WebGPU Shading Language (WGSL)---an apex-level shading language that transpiles to backend API languages---ensuring compatibility with all backend graphics APIs: Vulkan, Direct3D, and Metal. Our method incorporates a novel tile state representation that eliminates reliance on 64-bit atomics or memory barriers while maintaining correctness, along with a subgroup-size-agnostic intra-workgroup implementation that enables subgroup acceleration across all architectures.

@@

As shown in Figure~\ref{fig:execution_time}, performance relative to a no-omission baseline remains above 85\% even at near 100\% fallback rates across all devices---except for the Intel HD620, where only a 25\% fallback rate is induced. Notably, this performance level remains well above the equivalent \emph{RTS} performance of 66\%; \emph{DF} is consistently faster than the previous standard for portable scans. While Figures~\ref{fig:fallbacks_initiated} and~\ref{fig:total_spins} confirm that fallbacks are being induced, this test primarily measures the impact of increasing fallback operations rather than true blocking behavior. This is because workgroups are still able to proceed after an omission. In true starvation, a blocking workgroup stalls before posting its reduction, preventing subsequent workgroups from making progress. Unlike our simulation, which allows workgroups to eventually continue, true starvation results in increased lookback lengths, as blocked workgroups cannot complete \emph{Chained Propagation} and update their flags to inclusive. Furthermore, starvation effectively leads to \emph{occupancy reduction}: execution resources remain available, but fewer subgroups can context-switch to mitigate latency.
\section{Discussion}

@@

\subsection{Conclusion}
In this paper, we introduced \emph{Decoupled Fallback}, a single-pass \emph{Chained Scan} capable of achieving SOL performance across all progress models while remaining compatible with modern graphics APIs and hardware constraints. Our approach enables any workgroup to compute the reduction of any work tile, eliminating the risk of starvation. We reexamined the progress models proposed by Sorensen et al.~\cite{sorensen2021} in the context of dynamically assigned work, demonstrating that Chained Scan implementations require, at minimum, the OBE guarantee. Additionally, we highlighted the risks of running \emph{Decoupled Lookback} without FPG. Furthermore, we introduced a subgroup-accelerated yet subgroup-size-agnostic scan based on a generalized radix-$s$ Ladner-Fischer scan, along with a tile state representation that eliminates reliance on 64-bit atomics and memory barriers by distributing tile data across multiple values.

Our artifact is available in the following GitHub repository: TODO. The repository contains per API requirements and building instructions.
